---
title: "Quick look at dom species loss"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


#### Read in data + preprocessing (the end result of this chunk is a dataframe with 1 row for each plot/year/species)

```{r, message=F, warning=F}
library(tidyverse)
library(lme4)
library(patchwork)
library(emmeans)
library(here)


cover_april2020 <- read_csv(here("output-data", "NutNet_FullCoverData_ProcessedApril2020.csv")) %>%
  select(site_name, site_code, year_trt, trt, plot, Taxon, RelAbund_group, DI,
         totplotcover.yr.live, max_cover, live, Family, rel_abundance_year0)

comb_april2018 <- read_csv(here("input-data", "comb-by-plot-clim-soil-diversity-09-Apr-2018.csv")) %>%
  select(site_code, year_trt, plot, rich, live_mass, even)

# combine cover + comb data
# filter to only get sites with at least 5 treatment years.
# filter to also get rid of comp.pt site, since it has problems with pretreatment data
nutnetdf <- left_join(cover_april2020, comb_april2018, by = c("site_code", "plot", "year_trt")) %>%
  group_by(site_code) %>%
  filter(n_distinct(year_trt) >= 5, site_code != "comp.pt") %>%
  mutate(RelAbund_group = fct_relevel(as.factor(RelAbund_group), 'Subordinate', after = 1)) %>%
  ungroup() %>%
  mutate(rel_cover = (max_cover/totplotcover.yr.live )*100)

# pull out just the groupings
relAbund_groups <- nutnetdf %>%
  select(site_code, plot, Taxon, RelAbund_group) %>%
  distinct()

# Our data doesn't come with any rows where abundance is zero, 
# so we assume that unseen combinations of the data are zeroes:
# To be specific, for each site, we create every combination of (Plot, Year, Treatment) within that site (1)
# Then we take all the Taxon found in that site (at any point) and
# do a Cartesian product of (Plot, Year, Treatment) x Taxon. (2)
# Then we create a column called PA to represent the new rows we added. (3)
# Then we add back DI (since it had to be removed to do the groupings) (4)
nutnetdf_allspp <- nutnetdf %>%
  select(site_name, site_code, plot, year_trt, trt, Taxon, rel_cover) %>%
  group_by(site_name) %>%
  nest() %>% 
  mutate(spread_df = purrr::map(data, ~spread(., key = Taxon, value = rel_cover, fill = 0) %>% #(1)
                                  pivot_longer(-c("site_code", "plot", "year_trt", "trt"), 
                                               names_to = "Taxon", values_to = "rel_cover"))) %>% #(2)
  unnest(spread_df) %>%
  ungroup() %>%
  arrange(site_code, plot, Taxon, year_trt) %>%
  select(-data) %>%
  left_join(relAbund_groups, by = c("site_code", "plot", "Taxon"))
  

```

#### Explore the importance of dominant species by splitting up the plots into timeseries -- before and after the loss of dominant species.

Some stats: 

* There are 127 (out of ~ 1400) plots with at least one dominant species
* Out of these 127 plots, 36 lose a dominant species in at least one year. (loss = missing that year)

```{r}
# pull out only plots with dom species
dom_split <- nutnetdf_allspp %>%
  left_join(comb_april2018, by = c('site_code', 'plot', 'year_trt')) %>%
  # get rid of species that aren't there in year 0
  filter(!is.na(RelAbund_group)) %>%
  mutate(dom_lost_sp = RelAbund_group == 'Dominant' & rel_cover == 0) %>%
  group_by(site_code, plot) %>%
  mutate(has_dom = any(RelAbund_group == 'Dominant', na.rm = T),
         dom_lost_plot = any(dom_lost_sp)) %>%
  ungroup() %>%
  filter(has_dom)


# record all times of dom sp loss for each plot
dom_sp_lost_times <- dom_split %>%
  filter(dom_lost_sp) %>%
  select(site_code, plot, year_trt) %>%
  unite("siteplot", site_code, plot, remove = F)
  
# pull out the FIRST time of dom sp loss for each plot (doesn't look at whether the sp comes back)
dom_sp_firstloss_time <- dom_sp_lost_times %>%
  group_by(siteplot) %>%
  slice_min(year_trt) %>%
  ungroup() %>%
  rename(first_loss_yr = year_trt)

dom_split %>%
  filter(dom_lost_plot) %>%
  unite("siteplot", site_code, plot, remove = F) %>%
  ggplot(aes(year_trt, live_mass)) + 
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank()) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(~siteplot) +
  geom_vline(aes(xintercept = year_trt), data = dom_sp_lost_times)
```

#### Looking before/after first instance of dom sp loss

*NOTE: This is comparing live mass before and after the FIRST vertical line in each plot (see figure above). This is almost certainly not optimal*
```{r}
# create an indicator for before/after the first instance of dom sp loss in each plot
first_loss_df <- dom_split %>%
  inner_join(dom_sp_firstloss_time, by = c('site_code', 'plot')) %>%
  mutate(after_first_loss = year_trt > first_loss_yr) %>%
  filter(!is.na(live_mass))


# plot the data
first_loss_bar <- ggplot(first_loss_df, aes(after_first_loss, log(live_mass))) +
  geom_col() +
  labs(title = 'maybe there is a decrease in biomass after dom sp loss (?)')

first_loss_ts <- ggplot(first_loss_df, aes(year_trt, log(live_mass))) +
  geom_point(aes(color = after_first_loss)) +
  labs(title = "but.. maybe no difference after taking into account year_trt (?)")
  
first_loss_bar + first_loss_ts + 
  plot_layout(nrow = 2)
  
```

#### Model to control for trt/year/plot

Regression on live mass ~ "indicator for whether a dominant species was lost at some previous time point", using only the subset of the data with a dom species. Control for treatment and year, with varying intercept for site/plot 

Fit model to see if the difference between masses is significant, after including trt/year/site
```{r}
# fit model
first_loss_mass_mod <- lmer(log(live_mass) ~ after_first_loss + trt + year_trt 
                            + (1|site_code/plot), 
                 data = first_loss_df)

# see the predicted mean live mass between the two groups
# the means are very similar -- from this model, the difference is not significant.
emmeans(first_loss_mass_mod, ~after_first_loss)
```

To summarize: if we only look at plots with dominant species loss, there likely isn't a large difference in live biomass before and after the *first instance* of dominant species loss

A possible explanation is that I split the groups up incorrectly -- using the first instance of dominant species loss is probably a bad idea, esp when we see that dominant species often leave and come back to a plot... but what's a better way to do it?

Another possible explanation is that we need to consider more factors. What about the weather/soil conditions/ect?

